---
title: "FONDECYT Data Analysis"
author: "FONDECYT Research Project"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# FONDECYT Data Analysis

This document provides an interactive analysis of the FONDECYT research project data, extracted from the PostgreSQL database dump.

```{r libraries}
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(tidyr)
library(stringr)
library(knitr)
library(DT)
library(plotly)
```

## Data Import and Overview

```{r data-import}
# Load the extracted CSV data with complete metadata chain
cat("📂 Loading FONDECYT item responses data with complete metadata...\n")

# Look for the complete dataset (the only one generated by the updated script)
csv_file <- "data/item_responses_complete_202507271125.csv"

if (!file.exists(csv_file)) {
  stop("❌ Complete dataset not found. Please run the data extraction script:\n",
       "   Rscript data/extract_basic_responses.R\n",
       "This will generate the complete dataset with all metadata.")
}

# Load data
data <- read_csv(csv_file, show_col_types = FALSE)

# Basic data info
cat("✅ Complete enhanced dataset loaded successfully!\n")
cat("📊 Dataset shape:", nrow(data), "rows ×", ncol(data), "columns\n")
cat("💾 File size:", format(file.size(csv_file) / 1024^2, digits = 1), "MB\n")

# Verify we have complete metadata (all four levels)
has_item_details <- "item_label" %in% colnames(data)
has_admin_details <- "admin_start_datetime" %in% colnames(data)
has_subject_details <- "subject_id" %in% colnames(data)
has_followup_details <- "followup_session_number" %in% colnames(data)

if (has_item_details && has_admin_details && has_subject_details && has_followup_details) {
  cat("🔗 Complete metadata available: item + administration + subject + followup details\n")
} else {
  cat("⚠️  Unexpected metadata structure - some metadata levels may be missing\n")
}
```

## Column Documentation

```{r column-documentation}
# Create dynamic column documentation
base_descriptions <- list(
  "item_id" = "Unique identifier for the specific question/item being answered",
  "response_id" = "Unique identifier for each item response",
  "numeric_value" = "Numeric response value (for quantitative items)",
  "character_value" = "Text/character response value (for qualitative items)",
  "administration_id" = "Identifier linking to the assessment administration session",
  "response_time_options" = "Response time or timing-related options (possibly multiple choice timing)",
  "response_date" = "Date and time when the response was originally created",
  "response_updated" = "Date and time when the response was last modified/updated",
  "subject_value_id" = "Identifier for subject-specific value (may link to participant info)",
  "was_skipped" = "Boolean indicating if the item was skipped by the participant",
  "response_value" = "Unified response value (combines numeric and character responses)",
  "response_type" = "Type classification of the response (numeric/character)",
  "item_label" = "Short label/code identifying the specific item (e.g., PHQ_9.2_1)",
  "item_text" = "Full text of the question/item as presented to participants",
  "item_measure_id" = "Identifier for the measurement scale/instrument this item belongs to",
  "item_is_required" = "Boolean indicating if this item is required to be answered",
  "item_position" = "Position/order of this item within its measurement scale",
  "admin_start_datetime" = "Date and time when the administration session started",
  "admin_end_datetime" = "Date and time when the administration session ended",
  "admin_is_completed" = "Boolean indicating if the administration session was completed",
  "admin_original_administration_date" = "Original scheduled date for the administration",
  "subject_id" = "Unique identifier for the research subject who provided responses",
  "subject_type" = "Type/category of the research subject",
  "subject_study_site_id" = "Study site identifier where the subject participated",
  "follow_up_id" = "Identifier linking to followup session record",
  "followup_session_number" = "Sequential session number for the participant",
  "followup_attendance_status" = "Boolean indicating if participant attended the followup session",
  "followup_session_modality" = "Numeric code for session delivery modality (in-person, remote, etc.)",
  "followup_comments" = "Clinical notes and observations from the followup session",
  "followup_therapist" = "Name or identifier of the treating therapist"
)

# Build column info dynamically
column_info <- data.frame(
  Column = names(data),
  Type = sapply(data, function(x) class(x)[1]),
  Description = sapply(names(data), function(col) {
    if (col %in% names(base_descriptions)) {
      base_descriptions[[col]]
    } else {
      "Description not available"
    }
  }),
  Sample_Values = sapply(names(data), function(col) {
    sample_vals <- head(data[[col]], 3)
    # Truncate long text for readability
    sample_vals <- sapply(sample_vals, function(x) {
      if (is.na(x)) {
        "NA"
      } else if (is.character(x) && nchar(x) > 50) {
        paste0(substr(x, 1, 47), "...")
      } else {
        as.character(x)
      }
    })
    paste(sample_vals, collapse = ", ")
  }),
  stringsAsFactors = FALSE
)

# Display as interactive table
datatable(column_info, 
          options = list(pageLength = 20, scrollX = TRUE),
          caption = "Table 1: FONDECYT Dataset Column Documentation") %>%
  formatStyle(columns = 1:4, fontSize = '12px')
```

## Item Metadata Analysis

```{r item-metadata}
if (has_item_details) {
  cat("\n🔍 Item Metadata Analysis\n")
  cat("========================\n")
  
  # Count unique items
  unique_items <- data %>%
    select(item_id, item_label, item_text, item_measure_id, item_is_required, item_position) %>%
    distinct()
  
  cat("📊 Total unique items:", nrow(unique_items), "\n")
  cat("📊 Total measures/scales:", n_distinct(data$item_measure_id, na.rm = TRUE), "\n")
  cat("📊 Required items:", sum(unique_items$item_is_required, na.rm = TRUE), "\n")
  cat("📊 Optional items:", sum(!unique_items$item_is_required, na.rm = TRUE), "\n")
  
  # Sample of items by measure
  items_by_measure <- unique_items %>%
    count(item_measure_id, name = "items_count") %>%
    arrange(desc(items_count))
  
  cat("\n📋 Items per measure (top 10):\n")
  print(head(items_by_measure, 10))
  
  # Sample items table
  cat("\n📝 Sample items:\n")
  sample_items <- unique_items %>%
    slice_head(n = 10) %>%
    select(item_label, item_text, item_measure_id, item_is_required, item_position)
  
  datatable(sample_items,
            options = list(pageLength = 10, scrollX = TRUE),
            caption = "Sample of measurement items with metadata") %>%
    formatStyle(columns = 1:5, fontSize = '11px')
  
} else {
  cat("ℹ️  Item metadata not available in this dataset. Use the enhanced extraction to get item details.\n")
}
```

## Administration Metadata Analysis

```{r admin-metadata}
if (has_admin_details) {
  cat("\n🏛️ Administration Metadata Analysis\n")
  cat("==================================\n")
  
  # Count unique administrations
  unique_admins <- data %>%
    select(administration_id, admin_start_datetime, admin_end_datetime, admin_is_completed) %>%
    distinct()
  
  cat("📊 Total unique administrations:", nrow(unique_admins), "\n")
  cat("📊 Completed administrations:", sum(unique_admins$admin_is_completed, na.rm = TRUE), "\n")
  cat("📊 Incomplete administrations:", sum(!unique_admins$admin_is_completed, na.rm = TRUE), "\n")
  
  # Calculate completion rate
  completion_rate <- round(sum(unique_admins$admin_is_completed, na.rm = TRUE) / nrow(unique_admins) * 100, 1)
  cat("📊 Completion rate:", completion_rate, "%\n")
  
  # Duration analysis (for completed administrations)
  completed_admins <- unique_admins %>%
    filter(admin_is_completed == TRUE, !is.na(admin_start_datetime), !is.na(admin_end_datetime)) %>%
    mutate(
      start_time = as_datetime(admin_start_datetime),
      end_time = as_datetime(admin_end_datetime),
      duration_minutes = as.numeric(difftime(end_time, start_time, units = "mins"))
    ) %>%
    filter(duration_minutes > 0, duration_minutes < 1440)  # Between 0 and 24 hours
  
  if (nrow(completed_admins) > 0) {
    cat("\n⏱️  Administration Duration Summary (completed sessions):\n")
    cat("   Count:", nrow(completed_admins), "administrations\n")
    cat("   Median duration:", round(median(completed_admins$duration_minutes, na.rm = TRUE), 1), "minutes\n")
    cat("   Mean duration:", round(mean(completed_admins$duration_minutes, na.rm = TRUE), 1), "minutes\n")
    cat("   Range:", round(min(completed_admins$duration_minutes, na.rm = TRUE), 1), "-", 
        round(max(completed_admins$duration_minutes, na.rm = TRUE), 1), "minutes\n")
  }
  
  # Administration timeline
  admin_dates <- unique_admins %>%
    filter(!is.na(admin_start_datetime)) %>%
    mutate(admin_date = as_date(admin_start_datetime)) %>%
    count(admin_date, admin_is_completed) %>%
    arrange(admin_date)
  
  if (nrow(admin_dates) > 0) {
    cat("\n📅 Administration Timeline:\n")
    cat("   First administration:", min(admin_dates$admin_date), "\n")
    cat("   Last administration:", max(admin_dates$admin_date), "\n")
    
    # Show sample administrations
    cat("\n📋 Sample administrations:\n")
    sample_admins <- unique_admins %>%
      slice_head(n = 5) %>%
      select(administration_id, admin_start_datetime, admin_end_datetime, admin_is_completed)
    
    datatable(sample_admins,
              options = list(pageLength = 5, scrollX = TRUE),
              caption = "Sample administration sessions") %>%
      formatStyle(columns = 1:4, fontSize = '11px')
  }
  
} else {
  cat("ℹ️  Administration metadata not available in this dataset. Use the fully enhanced extraction to get administration details.\n")
}
```

## Subject Metadata Analysis

```{r subject-analysis, eval=has_subject_details}
if (has_subject_details) {
  cat("\n👥 Subject Metadata Analysis\n")
  cat("==========================\n")
  
  # Count unique subjects
  unique_subjects <- data %>%
    select(subject_id, subject_type, subject_study_site_id) %>%
    distinct()
  
  cat("📊 Total unique subjects:", nrow(unique_subjects), "\n")

  # Subject type distribution
  if ("subject_type" %in% colnames(data)) {
    subject_type_dist <- unique_subjects %>%
      count(subject_type, sort = TRUE) %>%
      mutate(percentage = round(n / sum(n) * 100, 1))
    
    cat("\n👤 Subject Type Distribution:\n")
    print(subject_type_dist)
  }

  # Study site distribution
  if ("subject_study_site_id" %in% colnames(data)) {
    site_dist <- unique_subjects %>%
      count(subject_study_site_id, sort = TRUE) %>%
      mutate(percentage = round(n / sum(n) * 100, 1))
    
    cat("\n🏥 Study Site Distribution:\n")
    print(site_dist)
  }

  # Responses per subject
  subject_response_counts <- data %>%
    count(subject_id, sort = TRUE)

  cat("\n📊 Subject Response Statistics:\n")
  cat("   Mean responses per subject:", round(mean(subject_response_counts$n), 1), "\n")
  cat("   Median responses per subject:", median(subject_response_counts$n), "\n")
  cat("   Range:", min(subject_response_counts$n), "-", max(subject_response_counts$n), "\n")

  # Top active subjects
  cat("\n🔝 Top 10 Most Active Subjects:\n")
  top_subjects <- head(subject_response_counts, 10)
  print(top_subjects)

  # Sample subject data
  cat("\n📋 Sample subject records:\n")
  sample_subjects <- unique_subjects %>%
    slice_head(n = 5)
  
  datatable(sample_subjects,
            options = list(pageLength = 5, scrollX = TRUE),
            caption = "Sample subject records") %>%
    formatStyle(columns = 1:ncol(sample_subjects), fontSize = '11px')
  
} else {
  cat("ℹ️  Subject metadata not available in this dataset. Use the complete enhanced extraction to get subject details.\n")
}
```

## Followup Session Analysis

```{r followup-analysis, eval=has_followup_details}
if (has_followup_details) {
  cat("\n🏥 Followup Session Analysis\n")
  cat("=========================\n")
  
  # Count responses with followup data
  responses_with_followup <- sum(!is.na(data$follow_up_id))
  followup_coverage <- round(responses_with_followup / nrow(data) * 100, 1)
  
  cat("📊 Followup data coverage:", responses_with_followup, "responses (", followup_coverage, "%)\n")
  
  # Unique followup sessions
  unique_followups <- data %>%
    filter(!is.na(follow_up_id)) %>%
    select(follow_up_id, followup_session_number, followup_attendance_status, followup_session_modality, followup_therapist) %>%
    distinct()
  
  cat("📊 Total unique followup sessions:", nrow(unique_followups), "\n")
  
  # Session number distribution
  if ("followup_session_number" %in% colnames(data)) {
    session_dist <- data %>%
      filter(!is.na(followup_session_number)) %>%
      count(followup_session_number, sort = TRUE)
    
    cat("\n📈 Session Number Distribution:\n")
    print(head(session_dist, 10))
    
    cat("\n   Session number range:", min(session_dist$followup_session_number, na.rm = TRUE), "-", 
        max(session_dist$followup_session_number, na.rm = TRUE), "\n")
  }
  
  # Attendance status
  if ("followup_attendance_status" %in% colnames(data)) {
    attendance_summary <- unique_followups %>%
      count(followup_attendance_status, name = "sessions") %>%
      mutate(percentage = round(sessions / sum(sessions) * 100, 1))
    
    cat("\n👥 Attendance Status:\n")
    print(attendance_summary)
  }
  
  # Session modality
  if ("followup_session_modality" %in% colnames(data)) {
    modality_summary <- unique_followups %>%
      filter(!is.na(followup_session_modality)) %>%
      count(followup_session_modality, name = "sessions") %>%
      mutate(percentage = round(sessions / sum(sessions) * 100, 1))
    
    cat("\n💻 Session Modality Distribution:\n")
    print(modality_summary)
  }
  
  # Therapist distribution
  if ("followup_therapist" %in% colnames(data)) {
    therapist_summary <- unique_followups %>%
      filter(!is.na(followup_therapist)) %>%
      count(followup_therapist, name = "sessions") %>%
      arrange(desc(sessions))
    
    cat("\n👩‍⚕️ Therapist Distribution:\n")
    cat("   Total therapists:", nrow(therapist_summary), "\n")
    cat("   Sessions per therapist range:", min(therapist_summary$sessions), "-", max(therapist_summary$sessions), "\n")
    
    cat("\n🔝 Top 10 Most Active Therapists:\n")
    print(head(therapist_summary, 10))
  }
  
  # Comments availability
  if ("followup_comments" %in% colnames(data)) {
    comments_count <- sum(!is.na(data$followup_comments) & data$followup_comments != "")
    comments_rate <- round(comments_count / responses_with_followup * 100, 1)
    
    cat("\n📝 Clinical Comments:\n")
    cat("   Sessions with comments:", comments_count, "out of", responses_with_followup, "(", comments_rate, "%)\n")
  }
  
  # Sample followup data
  cat("\n📋 Sample followup records:\n")
  sample_followups <- unique_followups %>%
    slice_head(n = 5)
  
  datatable(sample_followups,
            options = list(pageLength = 5, scrollX = TRUE),
            caption = "Sample followup session records") %>%
    formatStyle(columns = 1:ncol(sample_followups), fontSize = '11px')
  
} else {
  cat("ℹ️  Followup metadata not available in this dataset. Use the complete enhanced extraction to get followup details.\n")
}
```

## Data Summary Statistics

```{r data-summary}
# Summary statistics for key columns
cat("\n📈 Summary Statistics\n")
cat("====================\n")

# Date range
date_range <- data %>%
  mutate(
    response_date = as_datetime(response_date),
    response_updated = as_datetime(response_updated)
  ) %>%
  summarise(
    earliest_response = min(response_date, na.rm = TRUE),
    latest_response = max(response_date, na.rm = TRUE),
    earliest_update = min(response_updated, na.rm = TRUE),
    latest_update = max(response_updated, na.rm = TRUE),
    date_span_days = as.numeric(latest_response - earliest_response)
  )

cat("📅 Data collection period:\n")
cat("   From:", as.character(date_range$earliest_response), "\n")
cat("   To:  ", as.character(date_range$latest_response), "\n")
cat("   Span:", date_range$date_span_days, "days\n\n")

# Response type distribution
response_type_summary <- data %>%
  count(response_type, sort = TRUE) %>%
  mutate(percentage = round(n / sum(n) * 100, 1))

cat("📊 Response Types:\n")
print(response_type_summary)

# Skip rate
skip_rate <- data %>%
  summarise(
    total_responses = n(),
    skipped_responses = sum(was_skipped, na.rm = TRUE),
    skip_rate_percent = round(skipped_responses / total_responses * 100, 2)
  )

cat("\n⏭️  Skip Rate:", skip_rate$skip_rate_percent, "% (", skip_rate$skipped_responses, "out of", skip_rate$total_responses, ")\n")
```

## Dataset Preview

```{r data-preview}
# Show first few rows interactively
cat("👀 Dataset Preview (first 100 rows)\n")
cat("====================================\n")

# Convert dates for better display and handle potential issues
preview_data <- data %>%
  head(100) %>%
  mutate(
    response_date = as.character(response_date),
    response_updated = as.character(response_updated)
  )

# Interactive data table with simplified formatting
datatable(preview_data, 
          options = list(
            pageLength = 10, 
            scrollX = TRUE,
            scrollY = "400px",
            autoWidth = TRUE
          ),
          caption = "Table 2: FONDECYT Dataset Preview",
          filter = 'top',
          class = 'cell-border stripe hover') %>%
  formatStyle(columns = names(preview_data), fontSize = '11px')
```

## Data Dimensions and Structure

```{r data-dimensions}
# Unique counts for key identifiers
unique_counts <- data %>%
  summarise(
    unique_responses = n_distinct(response_id),
    unique_administrations = n_distinct(administration_id),
    unique_items = n_distinct(item_id),
    unique_subjects = n_distinct(subject_value_id, na.rm = TRUE),
    total_rows = n()
  )

cat("🔢 Dataset Dimensions\n")
cat("=====================\n")
cat("Total responses:        ", format(unique_counts$total_rows, big.mark = ","), "\n")
cat("Unique response IDs:    ", format(unique_counts$unique_responses, big.mark = ","), "\n")
cat("Unique administrations: ", format(unique_counts$unique_administrations, big.mark = ","), "\n")
cat("Unique items/questions: ", format(unique_counts$unique_items, big.mark = ","), "\n")
cat("Unique subjects:        ", format(unique_counts$unique_subjects, big.mark = ","), "\n")

# Check for data consistency
consistency_check <- unique_counts$total_rows == unique_counts$unique_responses
cat("\n✅ Data consistency: Each row represents a unique response:", consistency_check, "\n")
```
